Simple repression draft

### 2020-05-23 doc/appendix_01_bayesian_inference.tex
Ironed out merge conflict
Fixed my error in Fig 3D scaling, add MRM's new Fig4
Found an arithmetic error in log(k_i)~deps_P argument,
realized intercept has meaning after all. Updated plot
and text to reflect. Also added MRM's beautiful new Fig 4

% !TEX root = ./busty_transcription.tex
\section{Bayesian inference}\label{sec:bayesian}
\mmnote{
\begin{itemize}
\item Bayes theorem.
Elaborate on posterior as plausibility over
space of model parameters.
\item Choosing likelihoods.
Easy here, just the steady-state distributions of our master
equations, assuming iid single-cell measurements.
\item Choosing priors.
Not critical here because we are data-rich and nonhierarchical,
as long as we don't exclude ground truth. In other applications,
prior can matter a lot. This is a feature, not a bug: it's
revealing hidden structure that was implied by your modeling
assumptions and you just didn't realize it. Predictive checks to
verify you're not excluding any plausible values.
\item Marginalization/expectations.
Analytically hard, trivial with sampling.
\item MCMC.
Lot's of fun here, and deep analogies to stat mech and
Hamiltonian mechanics worth exploring. In a perfect world, you'd
wave a magic wand and generate independent draws from target
posterior. But we don't know how. Next best: generate a Markov
chain of $N$ \textit{correlated} samples, estimate an
autocorrelation time $\tau$, then you've got roughly $N/\tau$
effective samples. Obviously with some assumptions.
\item Posterior predictive checks.
So the posterior is fine, but is the model actually not wrong?
PPC is essential to verify that your model is a plausible
generative process for the data.
\item Discuss identifiability, degeneracy, and failed PPCs in main text. Better thru example anyway.
\end{itemize}
}

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig03.pdf}
\caption{\textbf{Constitutive promoter posterior inference and model
        comparison.}
        (A) The joint posterior density of model 5, the
        bursty promoter with negative binomially-distributed
        steady state, is plotted with MCMC samples.
        1D marginal densities are plotted as flanking histograms.
        The model was fit on \textit{lacUV5} data from~\cite{Jones2014}.
        (B) The ECDF of the observed population distribution of
        mRNA transcripts under the control of a constitutive
        lacUV5 promoter is shown in orange.
        The median posterior predictive ECDFs for models (1),
        Poisson, and (5), negative binomial, are plotted in
        green and blue, respectively.
        Lighter green and blue regions enclose 95\% of all
        posterior predictive samples from their respective models.
        Model (1) is in obvious contradiction with the data
        while model (5) is not.
        FISH data is again from~\cite{Jones2014}.
    (C) Joint posterior distributions for burst rate $k_i$ and mean burst size
    $b$ for 18 promoters from~\cite{Jones2014}. Each contour indicates the 95\%
    highest posterior density region for a particular promoter. Note that the
    vertical axis is shared with (D). Panel (D) plots the burst rate $k_i$ vs.\
    the binding energy for each promoter from energy matrices
    from~\cite{Brewster2012}.
    The dotted line shows the predicted slope according to
    $\ln k_i/\gamma \sim \text{constant} - \beta\Delta\epsilon_P$,
    described in text.}
\label{fig:constit_post_full}
\end{figure}

@@

How are the two coarse-grainings related? The only association that makes dimensional sense and produces the correct order-of-magnitude for the known parameters is to take
\begin{equation}
\frac{k_i}{\gamma} = \exp(-\beta\Delta\epsilon_P)
\label{eq:bursty_equil_corresp1}
\end{equation}
and
\begin{equation}
b = \frac{r}{\gamma}\frac{P}{N_{NS}}.
\label{eq:bursty_equil_corresp2}
\end{equation}
\fig{fig:constit_post_full}(D) shows that this linear scaling
between $\ln k_i$ and $-\beta\Delta\epsilon_P$ is approximately
true for all 18 constitutive promoters considered. The intercept
is an uninteresting function of the unknown model parameters, so
we only consider the slope of the scaling in
\fig{fig:constit_post_full}(D).
\mmnote{Wait, is that true or can we do something with the intercept, related to the burst size in some useful way? Needs more careful thought.}

While the associations represented by
\eq{eq:bursty_equil_corresp1} and \eq{eq:bursty_equil_corresp2}
appear to be born out by the scaling $\ln k_i \sim
-\beta\Delta\epsilon_P$, we do not find the association of
parameters encoded by
\eqrange{eq:bursty_equil_corresp1}{eq:bursty_equil_corresp2}
intuitive. If some deeper insight is waiting to be gleaned from it,
we leave it as an open problem.

@@

\subsection{Transcription factor kinetics can be inferred from FISH

>>>>

% !TEX root = ./busty_transcription.tex
\section{Bayesian inference}
\label{sec:bayesian}
\mmnote{
\begin{itemize}
\item Bayes theorem.
Elaborate on posterior as plausibility over
space of model parameters.
\item Choosing likelihoods.
Easy here, just the steady-state distributions of our master
equations, assuming iid single-cell measurements.
\item Choosing priors.
Not critical here because we are data-rich and nonhierarchical,
as long as we don't exclude ground truth. In other applications,
prior can matter a lot. This is a feature, not a bug: it's
revealing hidden structure that was implied by your modeling
assumptions and you just didn't realize it. Predictive checks to
verify you're not excluding any plausible values.
\item Marginalization/expectations.
Analytically hard, trivial with sampling.
\item MCMC.
Lot's of fun here, and deep analogies to stat mech and
Hamiltonian mechanics worth exploring. In a perfect world, you'd
wave a magic wand and generate independent draws from target
posterior. But we don't know how. Next best: generate a Markov
chain of $N$ \textit{correlated} samples, estimate an
autocorrelation time $\tau$, then you've got roughly $N/\tau$
effective samples. Obviously with some assumptions.
\item Posterior predictive checks.
So the posterior is fine, but is the model actually not wrong?
PPC is essential to verify that your model is a plausible
generative process for the data.
\item Discuss identifiability, degeneracy, and failed PPCs in main text. Better thru example anyway.
\end{itemize}
}

@@

@@

How are the two coarse-grainings related? The only association that makes dimensional sense and produces the correct order-of-magnitude for the known parameters is to take
\begin{equation}
\frac{k_i}{\gamma} = \exp(-\beta\Delta\epsilon_P)
\label{eq:bursty_equil_corresp1_appdx}
\end{equation}
and
\begin{equation}
b = \frac{r}{\gamma}\frac{P}{N_{NS}}.
\label{eq:bursty_equil_corresp2_appdx}
\end{equation}
\fig{fig:constit_post_full}(D) shows that this linear scaling
between $\ln k_i$ and $-\beta\Delta\epsilon_P$ is approximately
true for all 18 constitutive promoters considered. The intercept
is an uninteresting function of the unknown model parameters, so
we only consider the slope of the scaling in
\fig{fig:constit_post_full}(D).
\mmnote{Wait, is that true or can we do something with the intercept, related to the burst size in some useful way? Needs more careful thought.}

While the associations represented by
\eq{eq:bursty_equil_corresp1_appdx} and \eq{eq:bursty_equil_corresp2_appdx}
appear to be born out by the scaling $\ln k_i \sim
-\beta\Delta\epsilon_P$, we do not find the association of
parameters encoded by
\eqrange{eq:bursty_equil_corresp1_appdx}{eq:bursty_equil_corresp2_appdx}
intuitive. If some deeper insight is waiting to be gleaned from it,
we leave it as an open problem.

@@

\subsection{Transcription factor kinetics can be inferred from FISH}

### 2020-05-23 doc/section_04_bayesian_inference.tex
Fixed my error in Fig 3D scaling, add MRM's new Fig4
Found an arithmetic error in log(k_i)~deps_P argument,
realized intercept has meaning after all. Updated plot
and text to reflect. Also added MRM's beautiful new Fig 4

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig03.pdf}
\caption{\textbf{Constitutive promoter posterior inference and model
comparison.} (A) The joint posterior density of model 5, the bursty promoter
with negative binomially-distributed steady state, is plotted with MCMC samples.
1D marginal probability densities are plotted as flanking histograms. The model
was fit on \textit{lacUV5} data from~\cite{Jones2014}. (B) The empirical
distribution function (ECDF) of the observed population distribution of mRNA
transcripts under the control of a constitutive \textit{lacUV5} promoter is
shown in orange. The median posterior predictive ECDFs for models (1), Poisson,
and (5), negative binomial, are plotted in dark green and dark blue,
respectively. Lighter green and blue regions enclose 95\% of all posterior
predictive samples from their respective models. Model (1) is in obvious
contradiction with the data while model (5) is not. single-cell mRNA count data
is again from~\cite{Jones2014}. (C) Joint posterior distributions for burst rate
$k_i$ and mean burst size $b$ for 18 unregulated promoters
from~\cite{Jones2014}. Each contour indicates the 95\% highest posterior
probability density region for a particular promoter. Note that the vertical
axis is shared with (D). (D) Plots of the burst rate $k_i$ vs.\ the binding
energy for each promoter as predicted in~\cite{Brewster2012}. The dotted line
shows the predicted slope according to $\ln k_i/\gamma \sim \text{constant} -
\beta\Delta\varepsilon_P$, described in text. Each individual promoter is
labeled with a unique number in both (C) and (D) for cross comparison.}
\label{fig:constit_post_full}
\end{figure}

@@

Given this result, how are the two coarse-grainings related? The only
association that makes dimensional sense and produces the correct
order-of-magnitude for the known parameters is to take
\begin{equation}
\frac{k_i}{\gamma} = \exp(-\beta\Delta\varepsilon_P)
\label{eq:bursty_equil_corresp1}
\end{equation}
and
\begin{equation}
b = \frac{r}{\gamma}\frac{P}{N_{NS}}.
\label{eq:bursty_equil_corresp2}
\end{equation}
Figure~\ref{fig:constit_post_full}(D) shows that this linear scaling between
$\ln k_i$ and $-\beta\Delta\varepsilon_P$ is approximately true for all 18
constitutive promoters considered. The intercept is an uninteresting function of
the unknown model parameters, so we only consider the slope of the scaling in
\fig{fig:constit_post_full}(D).
\mmnote{Wait, is that true or can we do something with the intercept, related to the burst size in some useful way? Needs more careful thought.}

While the associations represented by
Eq.~\ref{eq:bursty_equil_corresp1} and Eq.~\ref{eq:bursty_equil_corresp2}
appear to be born out by the scaling $\ln k_i \sim -\beta\Delta\varepsilon_P$,
we do not find the association of parameters encoded by
\eqrange{eq:bursty_equil_corresp1}{eq:bursty_equil_corresp2}
intuitive. If some deeper insight is waiting to be gleaned from it, we leave it
as an open problem.

\mmnote{Outline, still to cover:
\begin{itemize}
\item All 18 promoters. Hey the model still works on things other than UV5,
cool. Interesting and somewhat counterintuitive scaling b/w burst rate and
binding energy. Puzzle in comparing w/ Chong2014 supercoiling model. Are we
missing important things?? Unclear, we leave as open question. Good enough for
our purposes: posterior is identifiable, PPC is great, and of the models we've
thought of it is unique in satisfying both.
\end{itemize}
}

\mmnote{Discuss puzzling comparison w/ Chong2014: if supercoiling is the thing, why are my burst sizes all the same but burst rates vary? And why is the duty cycle of my promoter so low, ie., bursts so short? COmpare their fig 7E; their $\beta/\alpha$ is my $k^+/k^-$. They have one or two genes with very small $\beta/\alpha$, does the \textit{galK} locus just happen to be that, or is there a deeper disagreement? Hard to say w/o more data.}

\subsection{Transcription factor kinetics can be inferred from FISH measurements}
\label{sec:rep_kinetics_inference}
\mmnote{Outline:
\begin{itemize}
\item Full model. Write Bayes, handwave 2F1 dist story in limits of weak,
strong, and intermediate repression.
\item Hey look it mostly works. With full model, 9D model is well identified,
though most of the individual experiments unsurprisingly are not: only the ratio
of repressor rates is identifiable. PPC is not perfect, it kinda misses some
rep/op pairs but overall it's surprisingly predictive considering how strong
were the assumptions that went into it.
\item Do we chalk this up to experimental imperfections or is there interesting
biology to be found in the disagreements? I'm inclined towards the former. Even
just catching cells in truly apples-to-apples growth phases, or matching up
parameters of imaging sessions weeks or months apart, is hardly trivial.
\item The alternative is to refine our model and/or add on complexity, and it's
not at all clear what additions to make to the model, especially how to produce
the long tail of repressed strains over UV5. I see no way to get that without
getting into complicated time-correlation/history effects, where
transcription...rebounds?? After being repressed?? Sounds odd.
\item Comparison between my inferred rates, equilibrium binding E, and
single-molecule measurements. Not quite as slam-dunk as I'd hoped because of
$\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good
really! With exception of one Oid data data point, free energies are w/in about
0.5 kT also. That's not massively larger than the repeatability threshold of
fold-change measurements, at maybe 0.1 to 0.3 kT.
\item Surprising that so many model parameters can be inferred from just a few
distributions that, by eye, don't have strong features! Does require some
assumptions, e.g., about equivalence of rates across experiments, but that seems
quite reasonable.
\end{itemize}
}

\fig{fig3:kR_inferences} presents inferred
transcription factor rates, compares them to
single-molecule measurements, and compares their ratios to gold-standard binding
energies.
\begin{figure}%[h!]
\centering
\includegraphics[width=0.9\textwidth]{../figures/fig3/fig3.ai}
\caption{\textbf{Simple repression model comparison.}
Panel (A) depicts the cartoon of simple repression for which we infer parameters.
\mmnote{Not sure panel (A) is necessary, we've already shown it in at least 2
other figs. We could instead show some alternative slices of 9D posterior as a
companion to (B), or something totally different\dots??}
(B) Contours which enclose 50\% and 95\% of the posterior
probability mass are shown for of several 2D slices of
the 9D posterior distribution, from fitting one
unbinding rate for each operator (O1, O2, O3) and one binding rate for each aTc
concentration (corresponding to an unknown mean repressor copy number). (C)
Ratios of our inferred unbinding rates are compared with operator binding energy
differences measured in Garcia \& Phillips 2011 (triangles) and Razo-Mejia et.\
al.\ 2018 (squares). (D) Unbinding rates inferred in this work are compared with
single-molecule measurements of the same from Hammar et.\ at.\ 2014
\mmnote{Johan Elf's group}. Values are of similar order of magnitude, but a
precise quantitative comparison unfortunately depends sensitively on the assumed
mRNA lifetime, which was not measured.}
\label{fig3:kR_inferences}
\end{figure}

\begin{figure}%[h!]
\centering
\includegraphics[width=0.9\textwidth]{../figures/figSIxx/ppc_many_pooled.pdf}
\caption{\textbf{Posterior predictive checks.}
(A) The posterior distribution from~\fig{fig3:kR_inferences}B is
shown again for context. (B), (C), and (D) Samples from the
posterior predictive distribution are shown for each of the datasets
used in the inference. Samples are sorted by operator and plotted separately
simply for visual clarity. The unrepressed promoter, UV5, is shown with each as
a reference point. \mmnote{Obviously the PPC is not perfect, but considering the
relative simplicity of the model, I'm pleasantly surprised how well it worked
and that the model was even identifiable at all. Interestingly, and perhaps not
surprisingly, only 3 of the 9 datasets were identifiable if fit in isolation:
otherwise, it was only by making assumptions that certain rates are equal across
datasets was identifiability achieved.}}
\label{fig4:ppc}
\end{figure}

>>>>

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig03.pdf}
\caption{\textbf{Constitutive promoter posterior inference and model
comparison.} (A) The joint posterior density of model 5, the bursty promoter
with negative binomially-distributed steady state, is plotted with MCMC samples.
1D marginal probability densities are plotted as flanking histograms. The model
was fit on \textit{lacUV5} data from~\cite{Jones2014}. (B) The empirical
distribution function (ECDF) of the observed population distribution of mRNA
transcripts under the control of a constitutive \textit{lacUV5} promoter is
shown in orange. The median posterior predictive ECDFs for models (1), Poisson,
and (5), negative binomial, are plotted in dark green and dark blue,
respectively. Lighter green and blue regions enclose 95\% of all posterior
predictive samples from their respective models. Model (1) is in obvious
contradiction with the data while model (5) is not. Single-cell mRNA count data
is again from~\cite{Jones2014}. (C) Joint posterior distributions for burst rate
$k_i$ and mean burst size $b$ for 18 unregulated promoters
from~\cite{Jones2014}. Each contour indicates the 95\% highest posterior
probability density region for a particular promoter. Note that the vertical
axis is shared with (D). (D) Plots of the burst rate $k_i$ vs.\ the binding
energy for each promoter as predicted in~\cite{Brewster2012}. The dotted line
shows the predicted slope according to Eq.~\ref{eq:bursty_equil_corresp1},
described in text. Each individual promoter is
labeled with a unique number in both (C) and (D) for cross comparison.}
\label{fig:constit_post_full}
\end{figure}

@@

Given this result, how are the two coarse-grainings related?
A quick estimate can shed some light.
Consider for instance the \textit{lacUV5} promoter, which we see from
Figure~\ref{fig:constit_post_full}(A) has $k_i/\gamma \sim b \sim \text{few}$,
from Figure~\ref{fig:constit_post_full}(B) has $\langle m \rangle \sim 20$,
and from~\cite{Brewster2012} has $\beta\Delta\epsilon_P \sim - 6.5$.
Further we generally assume $P/N_{NS} \sim 10^{-3}$ since
$N_{NS}\approx4.6\times10^6$ and $P\sim10^3$.
After some guess-and-check with these values, one finds the only
association that makes dimensional sense and produces the correct
order-of-magnitude for the known parameters is to take
\begin{equation}
\frac{k_i}{\gamma} = \frac{P}{N_{NS}} \exp(-\beta\Delta\varepsilon_P)
\label{eq:bursty_equil_corresp1}
\end{equation}
and
\begin{equation}
b = \frac{r}{\gamma}.
\label{eq:bursty_equil_corresp2}
\end{equation}
Figure~\ref{fig:constit_post_full}(D) shows that this linear scaling between
$\ln k_i$ and $-\beta\Delta\varepsilon_P$ is approximately true for all 18
constitutive promoters considered. The plotted line is simply
Eq.~\ref{eq:bursty_equil_corresp1} and assumes $P\approx 5000$.

While the associations represented by
Eq.~\ref{eq:bursty_equil_corresp1} and Eq.~\ref{eq:bursty_equil_corresp2}
appear to be born out by the data in Figure~\ref{fig:constit_post_full},
we do not find the association of parameters they imply to be intuitive.
We are also cautious to ascribe too much physical reality to the parameters.
Indeed, part of our point in comparing the various constitutive
promoter models is to demonstrate that these models each provide
an internally self-consistent framework that adequately describes
the data, but attempting to translate between models reveals
the dubious physical interpretation of their parameters.

\mmnote{Outline, still to cover:
\begin{itemize}
\item There is a puzzle in comparing w/ Chong2014 supercoiling
model: if supercoiling is the thing, why are my burst sizes all
the same but burst rates vary? And why is the duty cycle of my
promoter so low, ie., bursts so short? COmpare their fig 7E;
their $\beta/\alpha$ is my $k^+/k^-$. They have one or two genes
with very small $\beta/\alpha$, does the \textit{galK} locus just
happen to be that, or is there a deeper disagreement? Hard to say
w/o more data. Are we
missing important things?? Unclear, we leave as open question. Good enough for
our purposes: posterior is identifiable, PPC is great, and of the models we've
thought of it is unique in satisfying both.
\end{itemize}
}

\subsection{Transcription factor kinetics can be inferred from FISH measurements}
\label{sec:rep_kinetics_inference}

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig04.pdf}
\caption{\textbf{Simple repression parameter inference and comparison.}
(A) Contours which enclose 50\% and 95\% of the posterior
probability mass are shown for each of several 2D slices of
the 9D posterior distribution. The model assumes one
unbinding rate for each operator (O1, O2, O3) and one binding rate for each aTc
concentration (corresponding to an unknown mean repressor copy number).
(B, upper panel)
Ratios of our inferred unbinding rates are compared with
operator binding energy differences measured by
Garcia and Phillips~\cite{Garcia2011a} (triangles) and
Razo-Mejia et.\ al.~\cite{Razo-Mejia2018} (circles).
Blue glyphs compare O2-O1, while orange compare O1-Oid.
(B, lower panel)
Unbinding rates for O1 (cyan) and Oid (red) inferred in this work are compared with
single-molecule measurements of the same from Hammar et.\ at.\ 2014.
The comparison requires an assumption about mRNA lifetime
$\gamma^{-1}$, which was not measured, and we plot representative
choices assuming $\gamma^{-1}=3$~min (triangles) or $\gamma^{-1}=5$~min (squares).
(C) Posterior predictive checks are shown for each of the
datasets used in the inference of the model in (A). Observed FISH
data from~\cite{Jones2014} is plotted as black lines. The median
posterior predictive ECDF for each condition is plotted as a dark
colored line. Lighter colored bands enclose 95\% of all posterior
predictive samples for a given operator/repressor copy number pair.
Samples are sorted by operator and plotted separately for visual clarity.
The unrepressed promoter, UV5, is shown with each as a reference point.
}
\label{fig4:repressed_post_full}
\end{figure}

\mmnote{Outline:
\begin{itemize}
\item Full model. Write Bayes, handwave 2F1 dist story in limits of weak,
strong, and intermediate repression.
\item Hey look it mostly works. With full model, 9D model is well identified,
though most of the individual experiments unsurprisingly are not:
only 3 of the 9 datasets were identifiable if fit in isolation
and for the rest only the ratio of repressor rates is identifiable.
Only by making assumptions that certain rates
are equal across datasets was identifiability achieved.
PPC is not perfect, it kinda misses some
rep/op pairs but overall it's surprisingly predictive considering how strong
were the assumptions that went into it.
\item Do we chalk this up to experimental imperfections or is there interesting
biology to be found in the disagreements? I'm inclined towards the former. Even
just catching cells in truly apples-to-apples growth phases, or matching up
parameters of imaging sessions weeks or months apart, is hardly trivial.
\item The alternative is to refine our model and/or add on complexity, and it's
not at all clear what additions to make to the model, especially how to produce
the long tail of repressed strains over UV5. I see no way to get that without
getting into complicated time-correlation/history effects, where
transcription...rebounds?? After being repressed?? Sounds odd.
\item Comparison between my inferred rates, equilibrium binding E, and
single-molecule measurements. Not quite as slam-dunk as I'd hoped because of
$\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good
really! With exception of one Oid data data point, free energies are w/in about
0.5 kT also. That's not massively larger than the repeatability threshold of
fold-change measurements, at maybe 0.1 to 0.3 kT.
\item Surprising that so many model parameters can be inferred from just a few
distributions that, by eye, don't have strong features! Does require some
assumptions, e.g., about equivalence of rates across experiments, but that seems
quite reasonable.
\end{itemize}
}
